Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-16 paper/zhang-cook-laa-langrene-menendez.Rmd
writing in optim section, algo to appendix, small tweaks in vis diag section, a tweak of the fig 9 (WIP)

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis tool used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, characterised by the projection basis, that gives the most interesting low-dimensional view. The interestingness of the structure is defined by the index function, a function of the projection basis.

@@

Optimization methods can be divided into various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex] or stochastic optimisation [@nocedal2006numerical]. Our interest is on constrained optimization [@bertsekas2014constrained] as defined in the introduction section, and assuming it is not possible to find a solution to the problem in the way of a closed-form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathbb{A}$ space.

A large class of methods utilises the gradient information of the objective function to perform the optimisation iterations, with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. Derivative-free methods, which do not rely on the knowledge of the gradient, are more generally applicable. Derivative-free methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and heuristic methods [@sorensen2013metaheuristics]. Later, we will present a simulated annealing optimisation algorithm, which belongs to the class of random search methods, for optimisation with the guided tour.

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: Creeping random search (CRS), simulated annealing (SA), and pseudo derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of SA is summarised in Algorithm \ref{random-search}. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SAJO gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even the landing basis does have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} highlights how SAJO differs from SA in the inner loop.

Pseudo derivative (PD) search [@cook1995grand] uses a different strategy than SA and SAJO. Rather than randomly sample the basis space, PD first computes a search direction by evaluating vases close to the current basis. A step size is then chosen along the geodesic direction by another optimisation over an 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} summarises the inner loop of the PD.

@@

A data structure for diagnosing optimisers in projection pursuit guided tour is first defined. With a fixed data structure, the diagnostic plots presented later in this paper can be easily applied to diagnose other optimisation procedures.

@@

\begin{equation}
\renewcommand\arraystretch{2}  % default value: 1.0
\left[
\begin{array}{c|ccc|cc|cc}
t & \mathbf{A} & I & S & j &  l  & V_{1} & V_{2}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & V_{11} & V_{12}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1  & V_{21}  & V_{22}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2  & V_{31}  & V_{32}\\
\vdots & \vdots &\vdots &\vdots  &\vdots & \vdots &\vdots  &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_2 & \vdots  & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & 2& \vdots &  \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & k_2 &\vdots  & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
\hline
\vdots & \vdots & \vdots &\vdots  & J &  1 & \vdots & \vdots \\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_{J} & V_{T1}& V_{T2}\\
\hline
\vdots &\vdots & \vdots &\vdots & J  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & J & k_J &\vdots  & \vdots\\
\hline
\vdots& \vdots & \vdots & \vdots & J+1 & 1 & \vdots& \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T^\prime & \mathbf{A}_{T^\prime} & I_{T^\prime} &S_{T^\prime}  & J+1 &  l_{J+1} & V_{T^\prime 1}& V_{T^\prime 2}\\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\text{column name} \\
\hline
\text{search (start basis)} \\
\hline
\text{search} \\
\text{search} \\
\vdots \\
\text{search (accepted basis)} \\
\hline
\text{interpolate} \\
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\vdots \\
\hline
\text{search} \\
\vdots \\
\text{search (final basis)} \\
\hline
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\text{search (no output)} \\
\vdots \\
\text{search (no output)} \\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

The construction of diagnostic plots adopts the core concept of grammar of graphics [@wickham2010layered] in ggplot2. In grammar of graphics, plots are not produced by calling the commands, named by the appearance of the plot, i.e., boxplot and histogram, but by the concept of stacked layers. Seeing plots as stacked layers gives analysts the freedom to composite plots with multiple elements on hands.

## Checking how hard the optimiser is working

@@

## Examining the optimisation progress {#toy-interp}

@@

## Understanding the optimiser's coverage of the search space {#toy-pca}

@@

## Animating the diagnostic plots

@@

## The tour looking at itself

@@

## Forming a torus

@@

>>>>

The focus of this paper is on the optimisation problem arising in the projection pursuit guided tour [@buja2005computational], an exploratory data analysis tool used for detecting interesting structures in high-dimensional data through a set of lower-dimensional projections [@cook2008grand]. The goal of the optimisation is to identify the projection, characterised by the projection basis, that gives the most interesting low-dimensional view. A view is interesting if it can show some unexpected structure of the data, e.g., bi-modality, clustering and outliers.

@@

The optimisation problem of our interest is on constrained optimization [@bertsekas2014constrained] as defined in the introduction section, and assuming it is not possible to find a solution to the problem in the way of a closed-form. That is, the problem consists of finding the minimum or maximum of a function $f \in L^p$ in the constrained $\mathcal{A}$ space.

A large class of methods utilises the gradient information of the objective function, with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. As will be shown in the next section, the independent variable, in our optimisation problem, is a matrix of projection basis and the computational time required to perform differentation on a matrix could impede the render of tour animation. Hence, gradient-based methods are not the focus of this paper and considerations will be given to derivative-free methods. Derivative-free methods, which do not rely on the knowledge of the gradient, are more generally applicable. Derivative-free methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Below will introduce two random search methods: creeping random search and simulated annealing, and a pseudo derivative search.

<!-- Later, we will present a simulated annealing optimisation algorithm, which belongs to the class of random search methods, for optimisation with the guided tour. -->

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: Creeping random search (CRS), simulated annealing (SA), and pseudo derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of SA is summarised in Algorithm \ref{random-search} in the appendix. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

Simulated annealing (SA) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SAJO gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even the landing basis does have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} in the appendix highlights how SAJO differs from SA in the inner loop.

Pseudo derivative (PD) search [@cook1995grand] uses a different strategy than SA and SAJO. Rather than randomly sample the basis space, PD first computes a search direction by evaluating vases close to the current basis. A step size is then chosen along the geodesic direction by another optimisation over an 90 degree angle from $-\pi/4$ to $\pi/4$. The resulting candidate basis $\mathbf{A}_{**}$ is returned for the current iteration if it has a higher index value than the current one. Algorithm \ref{search-geodesic} in the appendix summarises the inner loop of the PD.

@@

A data structure for diagnosing optimisers in projection pursuit guided tour is first defined. With this data structure, four types of diagnostic plots are presented. In the section, examples will be given to show how these plot can be used to reveal issues in the projection pursuit optimisation.

@@

\begin{equation}
\renewcommand\arraystretch{2}  % default value: 1.0
\left[
\begin{array}{c|ccc|cc|cc}
t & \mathbf{A} & I & S & j &  l  & V_{1} & V_{2}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & V_{11} & V_{12}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1  & V_{21}  & V_{22}\\
\vdots & \vdots &\vdots &\vdots  &\vdots & \vdots &\vdots  &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_2 & \vdots  & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & k_2 &\vdots  & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
\hline
\vdots & \vdots & \vdots &\vdots  & J &  1 & \vdots & \vdots \\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_{J} & V_{T1}& V_{T2}\\
\hline
\vdots &\vdots & \vdots &\vdots & J  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & J & k_J &\vdots  & \vdots\\
\hline
\vdots& \vdots & \vdots & \vdots & J+1 & 1 & \vdots& \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T^\prime & \mathbf{A}_{T^\prime} & I_{T^\prime} &S_{T^\prime}  & J+1 &  l_{J+1} & V_{T^\prime 1}& V_{T^\prime 2}\\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\text{column name} \\
\hline
\text{search (start basis)} \\
\hline
\text{search} \\
\vdots \\
\text{search (accepted basis)} \\
\hline
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\vdots \\
\hline
\text{search} \\
\vdots \\
\text{search (final basis)} \\
\hline
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\text{search (no output)} \\
\vdots \\
\text{search (no output)} \\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

The construction of diagnostic plots adopts the core concept of grammar of graphics [@wickham2010layered] in ggplot2. In grammar of graphics, plots are not produced by calling the commands, named by the appearance of the plot, i.e., boxplot and histogram, but by the concept of stacked layers. Seeing plots as stacked layers gives analysts the freedom to compose plots with multiple elements on hand.

## Diagnostic 1: Checking how hard the optimiser is working

@@

## Diagnostic 2: Examining the optimisation progress {#toy-interp}

@@

## Diagnostic 3a: Understanding the optimiser's coverage of the search space {#toy-pca}

@@

## Diagnostic 3b: Animating the diagnostic plots

@@

## Diagnostic 4a: The tour looking at itself

@@

## Diagnostic 4b: Forming a torus

@@

\newpage

# Appendix

Three algorithms (Creeping random search, simulated annealing, and pseudo derivative) used in projection pursuit guided tour optimisation.

### 2021-03-16 paper/zhang-cook-laa-langrene-menendez.tex
render graphs
writing in optim section, algo to appendix, small tweaks in vis diag section, a tweak of the fig 9 (WIP)

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, characterised by the projection basis, that gives the most
interesting low-dimensional view. The interestingness of the structure
is defined by the index function, a function of the projection basis.

@@

Optimization methods can be divided into various classes, such as global
optimisation \citep{kelley1999iterative, fletcher2013practical}, convex
optimisation \citep{boyd2004convex} or stochastic optimisation
\citep{nocedal2006numerical}. Our interest is on constrained
optimization \citep{bertsekas2014constrained} as defined in the
introduction section, and assuming it is not possible to find a solution
to the problem in the way of a closed-form. That is, the problem
consists of finding the minimum or maximum of a function \(f \in L^p\)
in the constrained \(\mathbb{A}\) space.

A large class of methods utilises the gradient information of the
objective function to perform the optimisation iterations, with the most
notable one being the gradient ascent (descent) method. Although
gradient optimization methods are popular, they rely on the availability
of the objective function derivatives and on the complexity of the
constraints. Derivative-free methods, which do not rely on the knowledge
of the gradient, are more generally applicable. Derivative-free methods
have been developed over the years, where the emphasis is on finding, in
most cases, a near optimal solution. Examples of those include response
surface methodology \citep{box1951experimental}, stochastic
approximation \citep{robbins1951stochastic}, random search
\citep{fu2015handbook} and heuristic methods
\citep{sorensen2013metaheuristics}. Later, we will present a simulated
annealing optimisation algorithm, which belongs to the class of random
search methods, for optimisation with the guided tour.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. In projection pursuit, higher index values are
associated with interesting projections that depart from normality
\citep{diaconis1984asymptotics, huber1985projection}, e.g.~Legendre
index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial} and natural Hermite index
\citep{cook1993projection} for measuring the distance between the
density of projected data and a standard normal distribution, LDA index
\citep{lee2005projection} and PDA index \citep{lee2010projection} for
finding clustering structure in the data, and skewness-based index
\citep{Loperfido2018} and kurtosis-based index \citep{Loperfido2020} for
outlier detection. As a general visualisation method, tour produces
animations of high dimensional data via rotations of low-dimensional
planes. There are different tour implementations depending on how the
high dimensional space is investigated: grand tour \citep{cook2008grand}
selects the planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: Creeping random search (CRS), simulated annealing
(SA), and pseudo derivative (PD). Creeping random search (CRS) is a
random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search}.
\citet{posse1995projection} has proposed a slightly different cooling
scheme by introducing a halving parameter \(c\). In his proposal, the
\(\alpha\) is only adjusted if the last iteration takes more than \(c\)
times to find a better basis.

Simulated annealing (SA)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as CRS but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} highlights how SAJO differs from SA in the
inner loop.

Pseudo derivative (PD) search \citep{cook1995grand} uses a different
strategy than SA and SAJO. Rather than randomly sample the basis space,
PD first computes a search direction by evaluating vases close to the
current basis. A step size is then chosen along the geodesic direction
by another optimisation over an 90 degree angle from \(-\pi/4\) to
\(\pi/4\). The resulting candidate basis \(\mathbf{A}_{**}\) is returned
for the current iteration if it has a higher index value than the
current one. Algorithm \ref{search-geodesic} summarises the inner loop
of the PD.

@@

The construction of diagnostic plots adopts the core concept of grammar
of graphics \citep{wickham2010layered} in ggplot2. In grammar of
graphics, plots are not produced by calling the commands, named by the
appearance of the plot, i.e., boxplot and histogram, but by the concept
of stacked layers. Seeing plots as stacked layers gives analysts the
freedom to composite plots with multiple elements on hands.

\hypertarget{checking-how-hard-the-optimiser-is-working}{%
\subsection{Checking how hard the optimiser is
working}\label{checking-how-hard-the-optimiser-is-working}}

@@

\hypertarget{toy-interp}{%
\subsection{Examining the optimisation progress}\label{toy-interp}}

@@

\hypertarget{toy-pca}{%
\subsection{Understanding the optimiser's coverage of the search
space}\label{toy-pca}}

@@

\hypertarget{animating-the-diagnostic-plots}{%
\subsection{Animating the diagnostic
plots}\label{animating-the-diagnostic-plots}}

@@

\hypertarget{the-tour-looking-at-itself}{%
\subsection{The tour looking at
itself}\label{the-tour-looking-at-itself}}

@@

\hypertarget{forming-a-torus}{%
\subsection{Forming a torus}\label{forming-a-torus}}

@@

>>>>

The focus of this paper is on the optimisation problem arising in the
projection pursuit guided tour \citep{buja2005computational}, an
exploratory data analysis tool used for detecting interesting structures
in high-dimensional data through a set of lower-dimensional projections
\citep{cook2008grand}. The goal of the optimisation is to identify the
projection, characterised by the projection basis, that gives the most
interesting low-dimensional view. A view is interesting if it can show
some unexpected structure of the data, e.g., bi-modality, clustering and
outliers.

@@

The optimisation problem of our interest is on constrained optimization
\citep{bertsekas2014constrained} as defined in the introduction section,
and assuming it is not possible to find a solution to the problem in the
way of a closed-form. That is, the problem consists of finding the
minimum or maximum of a function \(f \in L^p\) in the constrained
\(\mathcal{A}\) space.

A large class of methods utilises the gradient information of the
objective function, with the most notable one being the gradient ascent
(descent) method. Although gradient optimization methods are popular,
they rely on the availability of the objective function derivatives and
on the complexity of the constraints. As will be shown in the next
section, the independent variable, in our optimisation problem, is a
matrix of projection basis and the computational time required to
perform differentation on a matrix could impede the render of tour
animation. Hence, gradient-based methods are not the focus of this paper
and considerations will be given to derivative-free methods.
Derivative-free methods, which do not rely on the knowledge of the
gradient, are more generally applicable. Derivative-free methods have
been developed over the years, where the emphasis is on finding, in most
cases, a near optimal solution. Below will introduce two random search
methods: creeping random search and simulated annealing, and a pseudo
derivative search.

@@

Let \(\mathbf{X}_{n \times p}\) be the data matrix with \(n\)
observations in \(p\) dimensions. A \(d\)-dimensional projection is a
linear transformation from \(\mathbb{R}^p\) into \(\mathbb{R}^d\)
defined as \(\mathbf{Y} = \mathbf{X} \cdot \mathbf{A}\), where
\(\mathbf{Y}_{n \times d}\) is the projected data and
\(\mathbf{A}_{p\times d}\) is the projection basis matrix. We define
\(f: \mathbb{R}^{n \times d} \mapsto \mathbb{R}\) to be an index
function that maps the projected data \(\mathbf{Y}\) onto a scalar
value. This is commonly known as the projection pursuit index function,
or just index function, and is used to measure the ``interestingness''
of a given projection. In projection pursuit, higher index values are
associated with interesting projections that depart from normality
\citep{diaconis1984asymptotics, huber1985projection}, e.g.~Legendre
index \citep{friedman1974projection}, Hermite index
\citep{hall1989polynomial} and natural Hermite index
\citep{cook1993projection} for measuring the distance between the
density of projected data and a standard normal distribution, LDA index
\citep{lee2005projection} and PDA index \citep{lee2010projection} for
finding clustering structure in the data, and skewness-based index
\citep{Loperfido2018} and kurtosis-based index \citep{Loperfido2020} for
outlier detection.

As a general visualisation method, tour produces animations of high
dimensional data via rotations of low-dimensional planes. There are
different tour implementations depending on how the high dimensional
space is investigated: grand tour \citep{cook2008grand} selects the
planes randomly to provide a general overview; manual tour
\citep{cook1997manual} gradually phases in and out one variable to
understand the contribution of that variable in the projection. Guided
tour, the main interest of this paper, chooses the planes with the aid
of projection pursuit, to gradually reveal the most interesting
projection. Given a random start, projection pursuit iteratively finds
bases with higher index values and the guided tour constructs a geodesic
interpolation between these planes to form a tour path. Figure
\ref{fig:tour-path} shows a sketch of the tour path where the blue
frames are produced by the projection pursuit optimisation and the white
frames are interpolations between the blue frames. Mathematical details
of the geodesic interpolation can be found in
\citet{buja2005computational}. The aforementioned tour method has been
implemented in the R package \CRANpkg{tourr} \citep{tourr}.

@@

Three optimisers have been implemented in the \CRANpkg{tourr}
\citep{tourr} package: Creeping random search (CRS), simulated annealing
(SA), and pseudo derivative (PD). Creeping random search (CRS) is a
random search optimiser that samples a candidate basis
\(\mathbf{A}_{l}\) in the neighbourhood of the current basis
\(\mathbf{A}_{\text{cur}}\) by
\(\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}\)
where \(\alpha\) controls the radius of the sampling neighbourhood and
\(\mathbf{A}_{\text{rand}}\) is generated randomly. \(\mathbf{A}_{l}\)
is then orthonormalised to fulfil the basis constraint. If
\(\mathbf{A}_{l}\) has an index value higher than the current basis
\(\mathbf{A}_{\text{cur}}\), the optimiser outputs \(\mathbf{A}_{l}\)
for guided tour to construct an interpolation path. The neighbourhood
parameter \(\alpha\) is adjusted by a cooling parameter:
\(\alpha_{j+1} = \alpha_j * \text{cooling}\) before the next iteration
starts. The optimiser terminates when the maximum number of iteration
\(l_{\max}\) is reached before a better basis can be found. The
algorithm of SA is summarised in Algorithm \ref{random-search} in the
appendix. \citet{posse1995projection} has proposed a slightly different
cooling scheme by introducing a halving parameter \(c\). In his
proposal, the \(\alpha\) is only adjusted if the last iteration takes
more than \(c\) times to find a better basis.

Simulated annealing (SA)
\citep{kirkpatrick1983optimization, bertsimas1993simulated} uses the
same sampling process as CRS but allows a probabilistic acceptance of a
basis with lower index value than the current. Given an initial value of
\(T_0\), the temperature at iteration \(l\) is defined as
\(T(l) = \frac{T_0}{\log(l + 1)}\). When a candidate basis fails to have
an index value larger than the current basis, SAJO gives it a second
chance to be accepted with probability
\[P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}\]
where \(I_{(\cdot)}\) denotes the index value of a given basis. This
implementation allows the optimiser to make a move and explore the basis
space even the landing basis does have a higher index value and hence
enable the optimiser to jump out of a local optimum. The algorithm
\ref{simulated-annealing} in the appendix highlights how SAJO differs
from SA in the inner loop.

Pseudo derivative (PD) search \citep{cook1995grand} uses a different
strategy than SA and SAJO. Rather than randomly sample the basis space,
PD first computes a search direction by evaluating vases close to the
current basis. A step size is then chosen along the geodesic direction
by another optimisation over an 90 degree angle from \(-\pi/4\) to
\(\pi/4\). The resulting candidate basis \(\mathbf{A}_{**}\) is returned
for the current iteration if it has a higher index value than the
current one. Algorithm \ref{search-geodesic} in the appendix summarises
the inner loop of the PD.

@@

The construction of diagnostic plots adopts the core concept of grammar
of graphics \citep{wickham2010layered} in ggplot2. In grammar of
graphics, plots are not produced by calling the commands, named by the
appearance of the plot, i.e., boxplot and histogram, but by the concept
of stacked layers. Seeing plots as stacked layers gives analysts the
freedom to compose plots with multiple elements on hand.

\hypertarget{diagnostic-1-checking-how-hard-the-optimiser-is-working}{%
\subsection{Diagnostic 1: Checking how hard the optimiser is
working}\label{diagnostic-1-checking-how-hard-the-optimiser-is-working}}

@@

\hypertarget{toy-interp}{%
\subsection{Diagnostic 2: Examining the optimisation
progress}\label{toy-interp}}

@@

\hypertarget{toy-pca}{%
\subsection{Diagnostic 3a: Understanding the optimiser's coverage of the
search space}\label{toy-pca}}

@@

\hypertarget{diagnostic-3b-animating-the-diagnostic-plots}{%
\subsection{Diagnostic 3b: Animating the diagnostic
plots}\label{diagnostic-3b-animating-the-diagnostic-plots}}

@@

\hypertarget{diagnostic-4a-the-tour-looking-at-itself}{%
\subsection{Diagnostic 4a: The tour looking at
itself}\label{diagnostic-4a-the-tour-looking-at-itself}}

@@

\hypertarget{diagnostic-4b-forming-a-torus}{%
\subsection{Diagnostic 4b: Forming a
torus}\label{diagnostic-4b-forming-a-torus}}

@@

\newpage

\hypertarget{appendix}{%
\section{Appendix}\label{appendix}}

Three algorithms (Creeping random search, simulated annealing, and
pseudo derivative) used in projection pursuit guided tour optimisation.
